Unifying Gompertzian growth with the communicable disease spreading paradigm

### 2022-05-15 main.tex
2nd readthrough + update small changes

A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the total population, $N \gg Y \ge I$, which seems well founded at the national level, viz.
\begin{align}
\label{eq:linSIR}
\frac{dY}{dt}& = \beta I\\
\frac{dI}{dt}& = (\beta - \alpha) I = \alpha (\phi(t) - 1) I,
\end{align}
and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.

@@

Some algebraic manipulations reveal that the diseased group is described by a single equation, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is assumed synonymous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$. 
When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model with a Gompertz model for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text. Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GompertzODE2param}), it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

@@

\section{Unifying the SIR model and the Gompertz model}
This remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

@@

One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compareLog})
\begin{equation}
\label{eq:networkSIR}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j \quad \forall i,
\end{equation}
using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice here that there is an implied causality going from the infected to susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{M}}{\partial x_i \partial x_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.

@@

Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''. 
\citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is quite noteworthy. It suggests that both systems share commonalities and means of communication.

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, Richard's parameter, $\theta$, in his growth model paradigm can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

@@

\subsection{Estimating model parameters}
The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig. \ref{fig:Gompertz_vs_SIRD}). Fitting is done with Python-Scipy with specified Jacobians and a 6-day windowed average of observed deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.

The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
\begin{equation}
\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\eta^2 + \text{Var}[\eta] \xi^2.
\end{equation}

\subsection{Code Availability}
The computer code required to produce the results are given in the following Github repository,
\begin{verbatim}
https://github.com/matzhaugen/GompertzManuscript
\end{verbatim}

@@

\subsection{A note on estimation}

@@

\begin{table}[H]
\label{tab:sird}
\centering
\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:SIRD_D}. Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
\begin{tabular}{llll}
\toprule
{} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
\midrule
Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
Austria        &     8(1) &    36(5) &    11(1) \\
Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
Sweden         &    19(5) &     8(2) &     3(1) \\
Italy          &   137(3) &  1.77(5) &  3.71(1) \\
France         &  240(10) &  1.24(7) &   8.1(4) \\
Spain          &   145(2) &  2.27(5) &  4.75(8) \\
Germany        &    61(4) &   3.9(3) &   6.5(5) \\
Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
US             &   312(5) &  0.76(2) &  1.83(7) \\
China          &  130(30) &   3.5(5) &    26(4) \\
\bottomrule
\end{tabular}

@@

\begin{table}[H]
\label{tab:gomp}
\centering
\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from \ref{eq:GompertzODE}.}
\begin{tabular}{llll}
\toprule
{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
\midrule
Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
France         &  10.2(1) &   8.2(1) &  11.0(8) \\
Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
US             &  7.53(7) &  5.32(7) &   8.4(5) \\
China          &   9.8(1) &   9.4(2) &    10(1) \\
\bottomrule
\end{tabular}

>>>>

A line of reasoning employed by \citet{Rypdal2020} to obtain the Gompertz curve is to linearize the SIR model by assuming
both the number of infected, $I(t)$, and cumulative infected, $Y(t)$, is much less than the initial pool of susceptible people, $N \gg Y \ge I$, which seems well founded at the national level, viz.
\begin{align}
\label{eq:linSIR}
\frac{dY}{dt}& = \beta I\\
\frac{dI}{dt}& = (\beta - \alpha) I = \alpha (\phi(t) - 1) I,
\end{align}
and where the number of recovered, $R(t)$, is under this linearization decoupled from the other variables.

@@

Some algebraic manipulations reveal that the diseased group is described by a single equation, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery, $\delta$ and $\alpha$. 
When only modeling the diseased however, knowledge of $N$ is irrelevant and the three parameters in (\ref{eq:SIRD_D}) are sufficient.

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of diseased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing a SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text. Although the Gompertz curve can also be obtained with a simple linear fit using (\ref{eq:GompertzODE2param}), it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with an inferred transmission rate in the SIRD model with 1 standard deviation in brackets, assuming bi-normally distributed parameter estimates. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

@@

\section{Unifying the SIR model and the Gompertz model}
The remarkable observation that the log-transformed domain is the natural one merits closer study. First, juxtapose the logistic model with the Gompertz model,

@@

One could add network interaction necessitating a corresponding matrix version of (\ref{eq:compareLog})
\begin{equation}
\label{eq:networkSIR}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j \quad \forall i,
\end{equation}
using the shorthanded $x_i=x_i(t)$ and with a fixed correlation governed by the network's growth rate, $\beta$, and adjacency matrix, $\{a_{i,j}\}=\mathbf{A} \in \mathbb{R}^{N \times N}$, a binary matrix with ones where the $i^{th}$ and $j^{th}$ nodes are connected, and zeroes otherwise. Notice that there is an implied causality going from the infected to susceptible, which will become relevant below. Furthermore, a linear correlation between variables is seen as the partial derivatives of the instantaneous growth rate with respect to pairs of microscopic variables,
\begin{equation}
\frac{\partial^2 \dot{M}}{\partial x_i \partial x_j} = -\frac{\beta}{N} a_{i,j}.
\end{equation}
Still, no Gompertz curve will emerge at the onset of the growth process. \citet{estrada2022networked} provide illuminating analysis on this topic.

@@

Thus, the emergence of the Gompertz curve at the macroscopic level suggests that the system is correlated, or coherent, presumably as a result of the simultaneous exposure to the same underlying stressor, but also due to the implied log-normal nature of the microscopic entities, where multiplication replaces addition as the aggregating operator \citep{zhang1994log}.
We can now further appreciate Richard's parametrization as a transition from non-collaborative to collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by \citet{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in (\ref{eq:rypdal}) as non-linear resource availability dependent on the overall magnitude, $Y(t)$, with growth at $\theta \rightarrow 0$ named ``maximally coherent''. 
\citet{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \citep{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by \citet{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

@@

\section{Conclusion}
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states, and that the perceived pathogen travels at infinite speed throughout the population, rejecting the possibility of a disease propagation stage through a perceived transmission network. In this vein, Richard's parameter, $\theta$, can be related to the number of higher order interactions with the susceptible and the infected in an SIR model \citep{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$.

@@

\subsection{Estimating model parameters}
The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (Fig. \ref{fig:Gompertz_vs_SIRD}). Fitting is done using Python-Scipy's non-linear least squares algorithm \verb|curve_fit| with specified Jacobians and a 6-day windowed average of observed deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.

The transmission rate is modeled as the product of $\eta \xi$ in (\ref{eq:SIRD_D}), and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \citep{nadarajah2016distribution},
\begin{equation}
\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\text{E}[\eta]^2 + \text{Var}[\eta] \text{E}[\xi]^2,
\end{equation}
where estimates are obtained by substituting the model parameters with those estimated through the technique outlined above.

\subsection{Code Availability}

Scripts used to produce the plots are downloadable in the form of a Jupyter notebook using Python from 
\newline
\verb|https://emf-research.fra1.digitaloceanspaces.com/gompertz/gompertz_vs_sird.ipynb|
% The computer code required to produce the results are given in the following Github repository,
% \begin{verbatim}
% https://github.com/matzhaugen/GompertzManuscript
% \end{verbatim}

@@

@@

\begin{table}[H]
\label{tab:sird}
\centering
\caption{Estimated values for the SIRD model fitted to various countries' observations as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq. (\ref{eq:SIRD_D}). Standard deviation estimates is given in the paranthesis corresponding to the last significant digit of the reported number.}
\begin{tabular}{llll}
\toprule
{} &           $\eta\times 10^{-1}$ &      $\xi \times 10^{4}$ &            $\kappa\times 10^{2}$ \\
\midrule
Belgium        &    94(8) &   3.2(2) &   9.3(7) \\
Portugal       &   6.6(4) &    35(2) &   4.5(3) \\
Austria        &     8(1) &    36(5) &    11(1) \\
Switzerland    &    22(1) &  13.6(7) &  11.0(5) \\
Sweden         &    19(5) &     8(2) &     3(1) \\
Italy          &   137(3) &  1.77(5) &  3.71(1) \\
France         &  240(10) &  1.24(7) &   8.1(4) \\
Spain          &   145(2) &  2.27(5) &  4.75(8) \\
Germany        &    61(4) &   3.9(3) &   6.5(5) \\
Denmark        &   2.7(2) &    96(8) &   4.3(3) \\
United Kingdom &   181(5) &  1.39(5) &   4.3(1) \\
Netherlands    &    29(1) &   8.9(5) &   4.3(2) \\
Turkey         &    34(3) &   6.8(5) &   6.9(6) \\
Iran           &  22.3(4) &  10.4(4) &  2.51(7) \\
US             &   312(5) &  0.76(2) &  1.83(7) \\
China          &  130(30) &   3.5(5) &    26(4) \\
\bottomrule
\end{tabular}

@@

\begin{table}[H]
\label{tab:gomp}
\centering
\caption{Estimated values for the Gompertz model fitted to various countries' observations, as shown in Figure \ref{fig:Gompertz_vs_SIRD_lnln_infty} using notation from eq. (\ref{eq:GompertzODE}).}
\begin{tabular}{llll}
\toprule
{} &             $\nu\times 10$ &           $\beta\times 10^{2}$ &        $\tilde{Y}\times 10^{2}$ \\
\midrule
Belgium        &   9.3(1) &   8.1(2) &  10.1(9) \\
Portugal       &  6.15(3) &  6.22(7) &   7.0(3) \\
Austria        &  7.28(7) &   8.1(2) &   8.1(7) \\
Switzerland    &  8.08(6) &   8.1(1) &   8.9(5) \\
Sweden         &   4.9(1) &   4.3(2) &     6(1) \\
Italy          &  7.60(6) &  5.88(7) &   8.4(4) \\
France         &  10.2(1) &   8.2(1) &  11.0(8) \\
Spain          &  9.41(9) &   7.5(1) &  10.2(7) \\
Germany        &  7.56(6) &  6.53(8) &   8.4(4) \\
Denmark        &  5.94(7) &   6.5(2) &   6.8(7) \\
United Kingdom &  8.46(5) &  6.54(7) &   9.3(4) \\
Netherlands    &  7.34(6) &  6.52(8) &   8.2(4) \\
Turkey         &  7.04(5) &  6.38(8) &   7.9(4) \\
Iran           &  5.87(6) &  5.04(8) &   6.7(4) \\
US             &  7.53(7) &  5.32(7) &   8.4(5) \\
China          &   9.8(1) &   9.4(2) &    10(1) \\
\bottomrule
\end{tabular}
